manuscript_analysis

### 2024-07-03 manuscript_analysis.Rmd
included multiple changes in the figures, also tried to figure out way to get derivatives of predicted model.

predictions_depth_temp(df_1)

# predictions(df_2)
# predictions(df_3)
# predictions(df_4)

@@

# Plot Figure 2. 
all_predictions <- model_predictions |>
  # exclude Lepomis gibbosus
  filter(species != "Lepomis_gibbosus") |> #lepomis gibbosus was excluded upon examination
  ggplot(aes(temp, fit)) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit), fill = factor(species)), alpha = 0.3) +
  geom_line(aes(color = factor(species))) +
  theme_bw() +
  facet_wrap(~species, scale = "free_y", labeller = labeller(species = label_wrap_gen(width = 10))) +
  theme(strip.background = element_rect(fill = "lightgrey"), strip.text = element_text(size = 6)) +
  scale_color_viridis(discrete = TRUE, guide = NULL, option = "G", aesthetics = c("color", "fill")) +
  xlab("Temperature") +
  ylab("Abundance")

all_predictions

# save the plot as TIFF
# tiff(paste("total_models/plots/all_predictions_models.tiff", sep = ""), units="in", width = 12, height=8, res=300)
# plot(all_predictions)
# 
# # Closing the graphical device
# dev.off()

@@

##Sensitivity testing
Testing if depth-driven temperature models can adequately fit abundance data compared to
a temperature and depth model. 
Models were compared in the eleven most abundant species.

@@

@@

# Plot Fig. 3
mycolor <- c("color" = "#A0E66A")

@@

# save as TIFF
# tiff(paste("total_models/plots/lm_species_richness_eutroph.tiff", sep = ""), units="in", width=12, height=5, res=300)
# 
# plot(figure_3)
# 
# # Closing the graphical device
# dev.off()

@@

plot_all_lms <- df_all_lms|> 
  ggplot(aes(species_richness, fit, color = category, fill = category)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), color = NA, alpha = 0.3) +
  theme_classic(base_size = 16) +
  ylab("Mean dissimilarity") +
  xlab("Species richness") +
  scale_fill_manual(values = mycolors,
                    aesthetics = c("color", "fill"), 
                    labels = c("endemic", "non-native", "non-endemic native", "translocated"),
                    name = NULL)
plot_all_lms
# save as TIFF
# tiff(paste("total_models/plots/lm_all.tiff", sep = ""), units="in", width=11, height=7, res=300)
# # plot(ggarrange(depth1, depth2, ncol = 2))
# # plot science discussion
# plot(plot_all_lms)
# 
# dev.off()
```

#### Fig. 4b 
Predictions of all species shown inside the categories. 
```{r, fig.show='hide'}

>>>>

# DO NOT RUN NOW, RUN LATER!
predictions_depth_temp(df_1)
predictions_depth_temp(df_2)
predictions_depth_temp(df_3)
# 4 not run yet, will wait with this
predictions_depth_temp(df_4)

# statistics of the gamms
total_models <- bind_rows(df_1, df_2, df_3, df_4)

# total_models |> 
#   filter(tot_abu > 20) |> 
#   distinct(Species, tot_abu)
stats_gamms(total_models)

@@

# read in the category table and rename categories
species_across_lakes <-  read_xlsx("species_category_across_lakes.xlsx") |> 
  select(-notes) |> 
    mutate(category = ifelse(category == "native", "non_endemic_native", category)) |>
  mutate(category = ifelse(category == "non_native_region", "non_endemic_native_and_translocated", category)) |> 
  mutate(category2 = factor("all"))

species_across_lakes$category <- as.factor(species_across_lakes$category)
levels(species_across_lakes$category)

#merge predictions with categories
predictions_categories <- merge(model_predictions, species_across_lakes)

# EVTL NEW PLOT 2
# change category names for facet_wrap
category_names <- c(
  `endemic` = "endemic",
  `endemic_and_translocated` = "translocated endemic",
  `non_endemic_native` = "non-endemic native",
  `non_endemic_native_and_translocated` = "translocated non-endemic native",
  `non_native` = "non-native"
)

# Rescale the predictions to 0-1 
predictions_categories_rescaled <- predictions_categories |>
  group_by(species) |>
  mutate(fit_rescaled = (fit - min(fit)) / (max(fit) - min(fit)))

plot_pred <- predictions_categories_rescaled |>
  ggplot(aes(temp, fit_rescaled, group = species, color = category)) +
  geom_line() +
  # how can I get the right color?
   # geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), alpha = 0.3) +
  theme_bw(base_size = 16) +
  ylab("Abundance") +
  xlab("Temperature") +
  scale_color_manual(values = mycolors1, guide = NULL) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        legend.position = "bottom",
        legend.title = element_blank())

# trying to group by category and color
predictions_categories |>
  ggplot(aes(temp, fit)) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit), fill = factor(category)), alpha = 0.3) +
  geom_line(aes(color = factor(category))) +
  theme_bw() +
  facet_wrap(~species + category, scale = "free_y", labeller = labeller(species = label_wrap_gen(width = 10))) +
  theme(strip.background = element_rect(fill = "lightgrey"), strip.text = element_text(size = 6)) +
  scale_color_manual(values = mycolors1, aesthetics = c("color", "fill")) +
  xlab("Temperature") +
  ylab("Abundance")

# OLD FIGURE 2

all_predictions <- predictions_categories |>
  ggplot(aes(temp, fit)) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit), fill = factor(species)), alpha = 0.3) +
  geom_line(aes(color = factor(species))) +
  theme_bw() +
  facet_wrap(~species, scale = "free_y", labeller = labeller(species = label_wrap_gen(width = 10))) +
  theme(strip.background = element_rect(fill = "lightgrey"), strip.text = element_text(size = 6)) +
  scale_color_viridis(discrete = TRUE, guide = NULL, option = "G", aesthetics = c("color", "fill")) +
  xlab("Temperature") +
  ylab("Abundance")

all_predictions
# save the plot as TIFF
# tiff(paste("total_models/plots/all_predictions_models.tiff", sep = ""), units="in", width = 12, height=8, res=300)
# plot(all_predictions)
# 
# # Closing the graphical device
# dev.off()

@@

##Sensitivity testing
NOT CHANGED 
Testing if depth-driven temperature models can adequately fit abundance data compared to
a temperature and depth model. 
Models were compared in the eleven most abundant species.

@@

# separating the dfs to separate model types
depth_binomial_gam <- df_binomial_gam |> 
  filter(tot_abu > 20)

temp_binomial_gam <- df_binomial_gam |> 
  filter(tot_abu < 20)

depth_abundance_gam <- df_abundance_gam |> 
  filter(tot_abu > 20)

temp_abundance_gam <- df_abundance_gam |> 
  filter(tot_abu < 20)

depth_binomial_re <- df_binomial_re |> 
  filter(tot_abu > 20)

temp_binomial_re <- df_binomial_re |> 
  filter(tot_abu < 20)

depth_abundance_re <- df_abundance_re |> 
  filter(tot_abu > 20)

temp_abundance_re <- df_abundance_re |> 
  filter(tot_abu < 20)

###Depth and temp models
```{r}
# DEPTH AND TEMPERATURE
# UNCLEAR HOW THIS CAN BE DONE
# Model 1: Binomial data without random intercepts

species_list <- depth_binomial_gam |>
  distinct(Species) |>
  pull(Species)

i <- "Salmo_sp_Blackspot"
j <- "Poschiavo"
# loop through the list of species
for (i in species_list) {
  data <- depth_binomial_gam |>
    filter(Species == i)

  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) +
                           s(Depth_sample, k = 3) + s(fProtocol, bs = 're'),
                         family = binomial)

    # using minimum and maximum temp in the lake for derivatives
     newdata <- tibble(mean_last_7days = seq(from = min(data_lake$temp, na.rm = TRUE),
                                             to = max(data_lake$temp, na.rm = TRUE)
                                             ,length = 200), Depth_sample =
                         mean(data$Depth_sample, na.rm = TRUE),
                       fProtocol = factor("VERT"))

     # calculate derivatives
     # derivatives <- derivatives(gam_output[[i]], data = newdata) |>
     #   mutate(fLake = factor(j)) |>
     #   mutate(species = factor(i))
     #   rename(temp = data)
     #

       test_derivatives <- derivative_samples(gam_output[[i]]
                                              , focal = "mean_last_7days", data = newdata)

       test_derivatives |>
         filter(.draw <= 200) |>
         ggplot(aes(x = mean_last_7days, y = .derivative, group = .draw)) +
         geom_line(alpha = 0.5)

       derivatives <- test_derivatives |> 
         filter(.draw == 500) |> 
         mutate(fLake = factor(j)) |>
         mutate(species = factor(i))

       derivatives |> 
         ggplot(aes(mean_last_7days, .derivative)) +
         geom_line()

# needs to be discussed

# nothing changed starting from here
############################################################################3##3
# Model 2.1: Abundance data and no random intercept for lake -> but binomial
# Coregonus_sp_benthic_profundal did not run with ZIP and was thus
# run with binomial.

species_list <- depth_abundance_gam |> 
  # binomial one
  filter(Species == "Coregonus_sp_benthic_profundal") |>
  distinct(Species) |> 
  pull(Species)

###Temp models
```{r}

@@

# Plot Fig. 3
"#78BE63"
mycolor <- c("color" = "#78BE63")
# mycolor <- c("color" = "#A0E66A")
"98D47C"

@@

figure_3
# save as TIFF
# tiff(paste("total_models/plots/lm_species_richness_eutroph.tiff", sep = ""), units="in", width=12, height=5, res=300)
# 
# plot(figure_3)
# 
# # Closing the graphical device
# dev.off()

###Divergence figure 3

```{r}
# first lm for species richness ~ maximum phosphorus

lm3 <- lm(mean_sign ~ sum_species + Phos_max, data = df_lm_excl)

# checking model assumptions
summary(lm3)
shapiro.test(resid(lm3))
lmtest::bptest(lm3)
plot(lm3)

# manually predict data
new_data3 <- tibble(sum_species = seq(from = min(df_lm_excl$sum_species), to = max(df_lm_excl$sum_species), length = 50), Phos_max = mean(df_lm_excl$Phos_max))

# predict lm
prediction_lm3 <- predict.lm(lm3, newdata = new_data3, se.fit = TRUE, type = "response")

# combine new data and predictions
bind_model3 <- cbind(new_data3, prediction_lm3) |> 
  mutate(category ="color")

# second lm for maximum phosphorus ~ species richness
lm4 <- lm(mean_sign ~ Phos_max + sum_species, data = df_lm_excl)

# checking model assumptions
summary(lm4)
shapiro.test(resid(lm4))
lmtest::bptest(lm4)
plot(lm4)

# new data
new_data4 <- tibble(Phos_max = seq(from = min(df_lm_excl$Phos_max), to = max(df_lm_excl$Phos_max),length = 50), sum_species = mean(df_lm_excl$sum_species))

# predict model
prediction_lm4 <- predict.lm(lm4, newdata = new_data4, se.fit = TRUE, type = "response")

# combine new data and predictions
bind_model4 <- cbind(new_data4, prediction_lm4) |> 
  mutate(category ="color")

# Plot Fig. 3
mycolor <- c("color" = "#88D44F")
# mycolor <- c("color" = "#A0E66A")
"98D47C"

lm3_plot <- bind_model3 |> 
  ggplot(aes(sum_species, fit,  color = category, fill = category)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
  theme_classic(base_size = 20) +
  ylab("Mean divergence") +
  xlab("Species richness") +
  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)

lm4_plot <- bind_model4 |> 
  ggplot(aes(Phos_max, fit, color = category, fill = category)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
  theme_classic(base_size = 20) +
  ylab("Mean divergence") +
  xlab("Maximum total phosphorus (µg/L)") +
  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)

# combine both plots
figure_divergence <- ggarrange(lm3_plot, lm4_plot)

figure_divergence
# save as TIFF
# tiff(paste("total_models/plots/lm_divergence_multiple_linear.tiff", sep = ""), units="in", width=12, height=5, res=300)
# 
# plot(figure_divergence)
# 
# # Closing the graphical device
# dev.off()

@@

plot_all_lms <- df_all_lms|> 
  ggplot(aes(species_richness, fit, color = category, fill = category, linetype = category)) +  
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), color = NA, alpha = 0.3) +
  theme_classic(base_size = 16) +
  ylab("Mean dissimilarity") +
  xlab("Species richness") +
  scale_fill_manual(values = mycolors,
                    aesthetics = c("color", "fill"),
                    labels = c("endemic", "non-native (ns)", "non-endemic native", "translocated (ns)")) +
   scale_linetype_manual(values = c("endemic" = "solid",
                                   "non_native" = "dashed",
                                   "non_endemic_native" = "solid",
                                   "translocated" = "dashed"), labels = c("endemic", "non-native (ns)", "non-endemic native", "translocated (ns)"))

plot_all_lms
# save as TIFF
# tiff(paste("total_models/plots/lm_all.tiff", sep = ""), units="in", width=11, height=7, res=300)
# # plot(ggarrange(depth1, depth2, ncol = 2))
# # plot science discussion
# plot(plot_all_lms)
# 
# dev.off()
```

####Divergence Fig. 4a

```{r}
# Lms for all categories: endemic, translcated, non_native and non_endemic_native
# with data on all lakes
# Endemic
deriv_lm1 <- lm(mean_sign ~ endemic, data = df_lm)

# checking model assumptions
summary(deriv_lm1)
shapiro.test(resid(deriv_lm1))
lmtest::bptest(deriv_lm1)
plot(deriv_lm1)

new_data_1 <- tibble(endemic = seq(from = min(df_lm$endemic), to = max(df_lm$endemic),
                                     length = 15))

prediction_endemic_1 <- predict.lm(deriv_lm1, newdata = new_data_1, se.fit = TRUE, type = "response")

df_deriv_endemic <- cbind(new_data_1, prediction_endemic_1) |> 
  mutate(category = factor("endemic")) |> 
  rename(species_richness = endemic)

deriv_lm2 <- lm(mean_sign ~ non_native, data = df_lm)

# checking model assumptions
summary(deriv_lm2)
shapiro.test(resid(deriv_lm2))
lmtest::bptest(deriv_lm2)
plot(deriv_lm2)

new_data_2 <- tibble(non_native = seq(from = min(df_lm$non_native), to = max(df_lm$non_native),
                                   length = 15))

prediction_2 <- predict.lm(deriv_lm2, newdata = new_data_2, se.fit = TRUE, type = "response")

df_deriv_nn <- cbind(new_data_2, prediction_2)  |> 
  mutate(category = factor("non_native")) |> 
  rename(species_richness = non_native)

deriv_lm3 <- lm(mean_sign ~ native, data = df_lm)

# checking model assumptions
summary(deriv_lm3)
shapiro.test(resid(deriv_lm3))
lmtest::bptest(deriv_lm3)
plot(deriv_lm3)

new_data_3 <- tibble(native = seq(from = min(df_lm$native), to = max(df_lm$native),
                                      length = 15))

prediction_3 <- predict.lm(deriv_lm3, newdata = new_data_3, se.fit = TRUE, type = "response")

df_deriv_nne <- cbind(new_data_3, prediction_3)  |> 
  mutate(category = factor("non_endemic_native")) |> 
  rename(species_richness = native)

deriv_lm4 <- lm(mean_sign ~ non_native_region, data = df_lm)

# checking model assumptions
summary(deriv_lm4)
shapiro.test(resid(deriv_lm4))
lmtest::bptest(deriv_lm4)
plot(deriv_lm4)

new_data_4 <- tibble(non_native_region = seq(from = min(df_lm$non_native_region), to = max(df_lm$non_native_region),
                                              length = 15))

prediction_4 <- predict.lm(deriv_lm4, newdata = new_data_4, se.fit = TRUE, type = "response")

df_deriv_trans <- cbind(new_data_4, prediction_4)  |> 
  mutate(category = factor("translocated")) |> 
  rename(species_richness = non_native_region)

# Combine all four lm predictions
df_deriv_lms <- rbind(df_deriv_endemic, df_deriv_nn, df_deriv_nne, df_deriv_trans)

lm_divergence_all <- df_deriv_lms|> 
  ggplot(aes(species_richness, fit, color = category, fill = category)) +
  geom_line(linewidth = 1, linetype = "dashed") +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), color = NA, alpha = 0.3) +
  theme_classic(base_size = 16) +
  ylab("Mean divergence") +
  xlab("Species richness") +
  scale_fill_manual(values = mycolors,
                    aesthetics = c("color", "fill"), 
                    labels = c("endemic", "non-native", "non-endemic native", "translocated"),
                    name = NULL)

lm_divergence_all

# save as TIFF
# tiff(paste("total_models/plots/lm_divergence_all.tiff", sep = ""), units="in", width=11, height=7, res=300)
# # plot(ggarrange(depth1, depth2, ncol = 2))
# # plot science discussion
# plot(lm_divergence_all)
# 
# dev.off()
```
#### Fig. 4b 
Predictions of all species shown inside the categories. 
```{r, fig.show='hide'}
